---
title: "HW3"
author: "Rishabh Khurana"
date: "11/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data <- read.csv("Diabetes.csv")
library(ggplot2)
data$gender <- as.factor(data$gender)
data$location <- as.factor(data$location)
data$frame <- as.factor(data$frame)
```

```{r}
plot1 <- ggplot(data = data, mapping = aes(gender, ratio)) + geom_boxplot()
data$whrat <- data$waist/data$hip
plot2 <- ggplot(data = data, mapping = aes(whrat, chol)) + geom_point()
plot3 <- ggplot(data = data, mapping = aes(height, weight)) + geom_point()
```

```{r}
plot1
```
Plot 1 shows the differences in the distributions of the cholesterol HDL ratio based on gender. Using the box plot, we can see that males have a more varied distribution as opposed to females, along with a lower mean overall. 

```{r}
plot2
```
Plot 2 shows the distribution between Waist/Hip ratio and cholesterol levels. The distribution follows a weak positive correlation, as the cholesterol levels slightly rise as the ratio increases. 

```{r}
plot3
```
Plot 3 shows how weight and height are correlated in diabetes patients. As we can see, height and weight have a weak correlation. Furthermore, the vast majority of patients weigh between 150 and 200, which creates some doubt how prevalent obesity is in diabetes patients
